---
title: "Data Science II Midterm"
author: "Arghya Kannadaguli (ak5357), Carolina Montes Garcia (cm3341), Fang Wang (fw2394)"
date: "2025-03-25"
output: html_document
---

## **Assignment Description**

A researcher conducted a study to evaluate antibody responses to a newly authorized vaccine. Understanding antibody responses can help inform vaccine effectiveness and identify populations with weaker immune responses. The antibody level measured via dry blood spot (log-transformed) is the primary outcome of interest. The researcher also collected data on potential predictors to understand factors influencing the antibody response. Here is a description of the dataset (dat1.RData): 

| Variable Name (colname) | Description |
| --- | --- |
| ID (id) | Participant ID |
| Age (age) | Age (in years) |
| Gender (gender) | 1 = Male, 0 = Female |
| Race/ethnicity (race) | 1 = White, 2 = Asian, 3 = Black, 4 =| Hispanic |
| Smoking (smoking) | 0 = Never smoked, 1 = Former smoker, 2 = Current| smoker |
| Height (height) | Height (in centimeters) |
| Weight (weight) | Weight (in kilograms) |
| BMI (bmi) | Body Mass Index; BMI = weight (kg) / [height (m)]² |
| Diabetes (diabetes) | 0 = No, 1 = Yes |
| Hypertension (hypertension) | 0 = No, 1 = Yes |
| Systolic blood pressure (SBP) | Systolic blood pressure (mmHg) |
| LDL cholesterol (LDL) | LDL cholesterol (mg/dL) |
| Time since vaccination (time) | Time since vaccination (in days) |
| Log-transformed antibody level (log_antibody) | Response variable |

Using this dataset, the researcher planned to build a prediction model of antibody levels, aiming to understand how demographic and clinical factors influence antibody responses and how antibody levels change over time following vaccination. Please help the researcher build this model. 

A few months after the initial data collection, the researcher collected a new, independent dataset (dat2.RData) with the same structure as dat1.RData. Evaluate the robustness and generalizability of your prediction model using the new dataset. Does the model generalize well to this new dataset? If the model does not generalize well, discuss possible reasons. 

## **Setup**

```{r setup, message = FALSE, warning = FALSE}
# LOAD LIBRARIES
library(tidyverse)
library(dplyr)
library(ggplot2)
library(pdp)
library(gt)
library(gtsummary)
library(ggpubr)
library (caret)
library(mgcv)
library(earth)
library(tidyverse)
library(rsample)

# KNIT SETTINGS
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  out.width = "90%",
  fig.align = "center"
)

# FIGURE SETTINGS
theme_set(
  theme(
    #legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, margin = margin(b = 5), face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, margin = margin(b = 10), color = "azure4", face = "bold", size = 8)
  )
)
```

##### **Load Data**
```{r}
load("data/dat1.RData")
load("data/dat2.RData")

dat1 = as_tibble(dat1) |> 
  janitor::clean_names() |> 
  mutate(across(c(gender, race, smoking, hypertension, diabetes), as.factor))

dat2 = as_tibble(dat2) |> 
  janitor::clean_names() |> 
  mutate(across(c(gender, race, smoking, hypertension, diabetes), as.factor))

dat1_decoded = dat1 |> 
  mutate(
    across(c(gender, race, smoking, hypertension, diabetes), as.character),
    across(c(gender, race, smoking, hypertension, diabetes), as.numeric),
    gender = factor(case_match(gender, 0 ~ "Female", 1 ~ "Male"), levels = c("Female", "Male")),
    race = factor(case_match(race, 1 ~ "White", 2 ~ "Asian", 3 ~ "Black", 4 ~ "Hispanic"), levels = c("White", "Asian", "Black", "Hispanic")),
    smoking = factor(case_match(smoking, 0 ~ "Never smoked", 1 ~ "Former smoker", 2 ~ "Current smoker"), levels = c("Never smoked", "Former smoker", "Current smoker")),
    hypertension = factor(case_match(hypertension, 0 ~ "No", 1 ~ "Yes"), levels = c("No", "Yes")),
    diabetes = factor(case_match(diabetes, 0 ~ "No", 1 ~ "Yes"), levels = c("No", "Yes"))
  )

head(dat1_decoded, 5) |> 
  knitr::kable(digits = 2)
```

## **General Exploratory Data Analysis (EDA)**

##### **Summary Statistics**
```{r message = FALSE}
dat1_decoded |> 
  select(-id) |> 
  tbl_summary(
    statistic = list (all_continuous() ~ "{min} - {max} | Mean: {mean} (SD: {sd})", 
                      all_categorical() ~ "{n} ({p}%)"),
    label = list(
      id = "ID (id)",
      age = "Age (age) ",
      gender = "Gender (gender)",
      race = "Race/ethnicity (race)",
      smoking = "Smoking (smoking)",
      height = "Height (height)",
      weight = "Weight (weight)",
      bmi = "BMI (bmi)",
      diabetes = "Diabetes (diabetes)",
      hypertension = "Hypertension (hypertension)",
      sbp = "Systolic blood pressure (sbp)",
      ldl = "LDL cholesterol (ldl)",
      time = "Time since vaccination (time)",
      log_antibody = "Log-transformed antibody level (log_antibody)"
    )
  ) |> 
  modify_caption("**Summary Statistics for Dataset 1**") #|> 
  # as_gt() |>
  # gtsave("figures/summary_table.png")
```

##### **Univariate Plots**

**Categorical Variable Barplots**
```{r}
p_gender = ggplot(dat1_decoded, aes(x = gender)) + 
  geom_bar(fill = "skyblue", color = "black", lwd = 0.5) + 
  labs(title = "Gender\nDistribution", x = "Gender", y = "Count")

p_race = ggplot(dat1_decoded, aes(x = race)) + 
  geom_bar(fill= "orange", color = "black", lwd = 0.5) + 
  labs(title = "Race/Ethnicity\nDistribution", x = "Race/Ethnicity", y = "Count")

p_smoking = dat1_decoded |> 
  mutate(smoking = fct_rev(as.factor(str_replace(smoking, " ", "\n")))) |> 
  ggplot(aes(x = smoking)) + 
  geom_bar(fill = "skyblue", color = "black", lwd = 0.5) + 
  labs(title = "Smoking Status\nDistribution", x = "Smoking", y = "Count")

p_diabetes = ggplot(dat1_decoded, aes(x = diabetes)) + 
  geom_bar(fill = "orange", color = "black", lwd = 0.5) + 
  labs(title = "Diabetes Status\nDistribution", x = "Diabetes", y = "Count")

p_hypertension = ggplot(dat1_decoded, aes(x = hypertension)) + 
  geom_bar(fill = "skyblue", color = "black", lwd = 0.5) + 
  labs(title = "Hypertension Status\nDistribution", x = "Hypertension", y = "Count")

barplot_combined = ggarrange(p_gender, p_race, p_smoking, p_diabetes, p_hypertension, nrow = 2, ncol = 3)

barplot_combined
```

**Numeric Variable Histograms**
```{r}
p_age = ggplot(dat1_decoded, aes(x = age)) + 
  geom_histogram(fill = "skyblue", color = "black", lwd = 0.5) + 
  labs(title = "Age\nDistribution", x = "Age", y = "Count")

p_height = ggplot(dat1_decoded, aes(x = height)) + 
  geom_histogram(fill = "orange", color = "black", lwd = 0.5) + 
  labs(title = "Height\nDistribution", x = "Height", y = "Count")

p_weight = ggplot(dat1_decoded, aes(x = weight)) + 
  geom_histogram(fill = "skyblue", color = "black", lwd = 0.5) + 
  labs(title = "Weight\nDistribution", x = "Weight", y = "Count")

p_bmi = ggplot(dat1_decoded, aes(x = bmi)) + 
  geom_histogram(fill = "orange", color = "black", lwd = 0.5) + 
  labs(title = "Body Mass Index\nDistribution", x = "BMI", y = "Count")

p_sbp = ggplot(dat1_decoded, aes(x = sbp)) + 
  geom_histogram(fill = "skyblue", color = "black", lwd = 0.5) + 
  labs(title = "Systolic Blood Pressure\nDistribution", x = "Systolic Blood Pressure", y = "Count")

p_ldl = ggplot(dat1_decoded, aes(x = ldl)) + 
  geom_histogram(fill = "orange", color = "black", lwd = 0.5) + 
  labs(title = "LDL Cholesterol\nDistribution", x = "LDL Cholesterol", y = "Count")

p_time = ggplot(dat1_decoded, aes(x = time)) + 
  geom_histogram(fill = "skyblue", color = "black", lwd = 0.5) + 
  labs(title = "Time Since Vax\nDistribution", x = "Time Since Vaccination", y = "Count")

histogram_combined = ggarrange(p_age, p_height, p_weight, p_bmi, p_sbp, p_ldl, p_time, nrow = 3, ncol = 3)

histogram_combined
```

## **Post-Split EDA** 

##### **Split Dataset**
```{r split}
# split dataset into training data and testing data:
set.seed(2025)
data_split = initial_split(dat1, prop =0.8)
training_data = training (data_split)
testing_data = testing (data_split)
# Cross-validation setup:
ctrl1 = trainControl(method = "cv", number = 10)
```

##### **Bivariate Plots**

**Density Plots of Response Variable by Categorical Variables**
```{r message = FALSE}
# Create function for plot generation
cat_plot = function(var, label, df = dat1_decoded){
  output_plot = df |> 
    ggplot(aes(x = log_antibody, fill = !!sym(var))) +
    geom_density(alpha = 0.5, lwd = 0.5) +
    labs(
      title = paste0("Log Antibody vs. ", label),
      x = "Log Antibody",
      y = "Density",
      fill = label
    )
  
  return(output_plot)
}

training_data_decoded = dat1_decoded |> 
  filter(id %in% training_data$id)

# Generate Plots
gender_plot = cat_plot("gender", "Gender", training_data_decoded)
race_plot = cat_plot("race", "Race", training_data_decoded)
smoking_plot = cat_plot("smoking", "Smoking Status", training_data_decoded)
hypertension_plot = cat_plot("hypertension", "Hypertension", training_data_decoded)
diabetes_plot = cat_plot("diabetes", "Diabetes", training_data_decoded)

# Display multi-plot figure (manually exporting to save multi-plot figure)
cat_vars_plot = ggarrange(gender_plot, race_plot, smoking_plot, hypertension_plot, diabetes_plot, nrow = 3, ncol = 2)
cat_vars_plot
```

**Trellis Plots for Numeric Variables**
```{r}
theme1 = trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)
featurePlot(x = training_data[, c("age", "height", "weight", "bmi", "sbp", "ldl", "time")], training_data$log_antibody, plot = "scatter", labels = c("", "Y"),
type = c("p"), layout = c(4, 2)) #ncol, nrow
```

**Scatterplots for Response with Numeric Variables**
```{r}
# Create function for plot generation
num_plot = function(var, label, xlab, df = dat1){
  output_plot = df |> 
    ggplot(aes(x = !!sym(var), y = log_antibody)) +
    geom_point(size = 0.3) +
    geom_smooth(lwd = 0.5, method = "loess", color = "red") +
    labs(
      title = paste0("Log Antibody vs. ", label),
      x = xlab,
      y = "Log Antibody"
    )
  
  return(output_plot)
}

# Generate Plots
age_plot = num_plot("age", "Age", "Age (years)", training_data)
height_plot = num_plot("height", "Height", "Height", training_data)
weight_plot = num_plot("weight", "Weight", "Weight", training_data)
bmi_plot = num_plot("bmi", "BMI", "Body Mass Index", training_data)
sbp_plot = num_plot("sbp", "SBP", "Systolic Blood Pressure", training_data)
ldl_plot = num_plot("ldl", "LDL", "LDL Cholesterol", training_data)
time_plot = num_plot("time", "Time", "Time Since Vaccination", training_data)

# Display multi-plot figure (manually exporting to save multi-plot figure)
cont_vars_plot = ggarrange(age_plot, height_plot, weight_plot, bmi_plot, sbp_plot, ldl_plot, time_plot, nrow = 3, ncol= 3)
cont_vars_plot
```

##### **Correlation Plot**
```{r}
dat1_x_matrix = model.matrix(log_antibody ~ ., data = dat1)[, -1]
corrplot::corrplot(cor(dat1_x_matrix), method = "circle", type = "full")
```

## **Model Training**

##### **Create Model Matrix**
```{r matrix}
# training_data
x = model.matrix(log_antibody~ .- id, training_data)[,-1] #remove the first column
y = training_data$log_antibody
```

##### **1: Basic OLS Model**

Train model.
```{r}
set.seed(2025)
lm_fit = train(
  x = x, 
  y = y,
  method = "lm",
  trControl = ctrl1
)

summary(lm_fit)
```

Visualize model.
```{r}
par(mfrow = c(2, 2))  # Arrange plots in 2x2 layout
plot(lm_fit$finalModel)  # Diagnostic plots
```

##### **2: Elastic Net**

Train model.
```{r}
set.seed(2025)
enet_fit = train(
    x = x,
    y = y,
    method = "glmnet",
    tuneGrid = expand.grid(
        alpha = seq(0, 1, length = 21),
        lambda = exp(seq(9, -3, length = 100))
    ),
    trControl = ctrl1
)

summary(enet_fit)
```

Optimal tuning parameter.
```{r}
enet_fit$bestTune
```

Visualize training process.
```{r}
par(mfrow = c(1, 1)) # reset plotting grid afterwards
plot(enet_fit, xTrans = log)
```

##### **3: Partial Least Squares**

```{r}
set.seed(2025)
pls_fit = train(x, y,
  method = "pls",
  tuneGrid = data.frame(ncomp = 1:14),
  trControl = ctrl1,
  preProcess = c("center", "scale"))

summary(pls_fit)
```

Optimal tuning parameter.
```{r}
pls_fit$bestTune$ncomp #best_ncomp
```

Visualize training process.
```{r}
ggplot(pls_fit, highlight = TRUE) + theme_bw()
```

##### **4: Generalized Additive Model (GAM)**

Train Model.
```{r gam_caret}
set.seed(2025)
gam_fit = train(
  log_antibody ~ age + height + weight + bmi + sbp + ldl + time + gender + race + smoking + diabetes + hypertension,
  data = training_data,
  method = "gam",
  trControl = ctrl1
)

summary(gam_fit)
```

Optimal model.
```{r}
model_gam = gam_fit$finalModel
```

Visualize model tuning.
```{r}
plot(model_gam, pages = 1, shade = TRUE, seWithMean = TRUE)
```

# **5: MARS**
```{r mars}
set.seed(2025)
mars_grid = expand.grid(degree = 1:4, nprune = 2:20)
mars_fit = train(x, y,
                  method = "earth",
                  tuneGrid = mars_grid,
                  trControl = ctrl1)

summary(mars_fit)
```

Optimal tuning parameters.
```{r}
mars_fit$bestTune # nprune: 8; degree: 1
```

Optimal model.
```{r}
coef(mars_fit$finalModel)
```

Visualize.
```{r}
ggplot(mars_fit)
```

```{r}
p1 = pdp::partial(mars_fit, pred.var = c("age"), grid.resolution = 10) |> autoplot()
p2 = pdp::partial(mars_fit, pred.var = c("bmi"), grid.resolution = 10) |> autoplot()
p3 = pdp::partial(mars_fit, pred.var = c("time"), grid.resolution = 10) |> autoplot()

gridExtra::grid.arrange(p1, p2, p3, ncol =3)
```

## **Model Selection**

##### **Comparing Models**

```{r}
resamp = resamples(list(OLS = lm_fit,
                        ENET = enet_fit,
                        PLS = pls_fit,
                        GAM = gam_fit, 
                        MARS = mars_fit))
summary(resamp)

bwplot(resamp, metric = "RMSE")
```

##### **GAM Model performs best.**

## **Estimating Generalizability**

```{r}
test_x = model.matrix(log_antibody ~ ., testing_data)[, -1]
gam_pred_dat1 = predict(gam_fit, newdata = testing_data)
gam_rmse_dat1 = sqrt(mean((gam_pred_dat1 - testing_data$log_antibody)^2))

gam_rmse_dat1
```

## **Assessing Generalizability**

```{r}
dat2_x = model.matrix(log_antibody ~ ., dat2)[, -1]
gam_pred_dat2 = predict(gam_fit, newdata = dat2)
gam_rmse_dat2 = sqrt(mean((gam_pred_dat2 - dat2$log_antibody)^2))

gam_rmse_dat2
```
