---
title: "midterm_Fang"
author: "Fang Wang"
date: "2025-03-29"
output: html_document
editor_options: 
  chunk_output_type: console
---
# Library
```{r library}
library (caret)
library(mgcv) 
library(pdp)
library(earth)
library(tidyverse)
library(rsample)
setwd("/Users/fangwang/Library/CloudStorage/OneDrive-ColumbiaUniversityIrvingMedicalCenter/Master's_program/Data Science II/Midterm Project")
```

# Data
```{r data}
load("dat1.RData") 
dat1 = dat1 |> 
  janitor::clean_names()
load("dat2.RData")
dat2 = dat2 |> 
  janitor::clean_names()
```

# EDA
```{r eda}
# Explore if predictors are continuous or categorical variables:
sapply(dat1, class)
# continuous variables:
## "age"
## "height"
## "weight"
## "bmi"
## "sbp"
## "ldl"
## "time"

# categorical varaibles:
## gender
## race
## smoking
## diabetes
## hypertension
```

# Split dataset
```{r split}
# split dataset into training data and testing data:
set.seed(2025)
data_split = initial_split(dat1, prop =0.8)
training_data = training (data_split)
testing_data = testing (data_split)
# Cross-validation setup:
ctrl1 = trainControl(method = "repeatedcv",
number = 10,
repeats = 5)
```

# Create Matrix
```{r matrix}
# training_data
x = model.matrix(log_antibody~ .- id, training_data)[,-1] #remove the first column
y = training_data$log_antibody
```

# EDA (continuous variables)
```{r eda}
# Get the image parameters of the current trellis graph
theme1 = trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)
featurePlot(x = training_data[, c("age", "height", "weight", "bmi", "sbp", "ldl", "time")], y, plot = "scatter", labels = c("", "Y"),
type = c("p"), layout = c(4, 2)) #ncol, nrow

ggsave(filename = "Trellis plot for continuous variables.png", plot = last_plot())
```
*All scatterplots look like circular clouds — no visible upward or downward trend. No strong linear relationship between these variables and log_antibody.Possibly no clear non-linear relationship either — at least from this visualization. The dots are symmetrically spread around a horizontal center — that’s usually a sign of no obvious association.*

# Correlation Matrix Plot
```{r}
corrplot::corrplot(cor(x), method = "circle", type = "full")
```
*"height", "weight", and "bmi" are strongly positively correlated with each other — large dark blue circles. "sbp" and "hypertension": also positively correlated. Strong correlations (e.g., weight and bmi) might signal multicollinearity — be careful when using these together in regression models.*

# Elastic Net
```{r enet}
enet_fit = train(
  x = x, 
  y = y,
  method = "glmnet",
  tuneGrid = expand.grid(
    alpha = seq(0, 1, length = 21),
    lambda = exp(seq(2, -3, length = 100))
  ),
  trControl = ctrl1
)
```


# PLS
```{r pls}
set.seed(2025)
pls_fit = train(x, y,
method = "pls",
tuneGrid = data.frame(ncomp = 1:14),
trControl = ctrl1,
preProcess = c("center", "scale"))
summary(pls_fit)

ggplot(pls_fit, highlight = TRUE) + theme_bw()

pls_fit$bestTune$ncomp #best_ncomp
```

# GAM
```{r gam_caret}
set.seed(2025)
gam_fit = train(
  log_antibody ~ age + height + weight + bmi + sbp + ldl + time + gender + race + smoking + diabetes + hypertension,
  data = training_data,
  method = "gam",
  trControl = ctrl1
)
summary(gam_fit)

model_gam = gam_fit$finalModel

plot(model_gam, pages = 1, shade = TRUE, seWithMean = TRUE)
```
*This model output is from a Generalized Additive Model (GAM), where both parametric terms (like gender, race, smoking) and nonlinear smooth terms (like age, BMI, etc.) are used to predict the outcome variable.*
*Linear effects: gender, race, smoking status, diabetes, hypertension*
*Nonlinear effects: age, systolic BP (sbp), LDL, BMI, time, height, weight, all modeled with smooth functions (s())*
*age, bmi and time have a strong nonlinear relationship with outcome.*

# MARS
```{r mars}
set.seed(2025)
mars_grid = expand.grid(degree = 1:4, nprune = 2:20)
mars_fit = train(x, y,
                  method = "earth",
                  tuneGrid = mars_grid,
                  trControl = ctrl1)
ggplot(mars_fit)
mars_fit$bestTune # nprune: 8; degree: 1
coef(mars_fit$finalModel)
```
*The simplest model (degree = 1) with around 8 terms performs best — lowest RMSE, no overfitting.*

*(Intercept) = 10.36:	Baseline value of the outcome when all other terms are 0.*
*gender = -0.29:	Being male (if coded as 1) reduces the outcome by 0.29 units compared to female (0).*
*h(bmi - 26.6) = -0.072:	If BMI > 26.6, the outcome decreases by 0.072 per unit above 26.6. Otherwise, this term is 0.*
*h(time - 57) = -0.0023:	If time > 57, outcome decreases slightly per unit above 57.*
*h(57 - time) = -0.034:	If time < 57, outcome decreases per unit below 57.*
*h(70 - age) = 0.02:	If age < 70, outcome increases per unit below 70.*
*smoking2 = -0.19:	Belonging to smoking group 2 decreases outcome by 0.20 units.*


# PLP
```{r plp}
p1 = pdp::partial(mars_fit, pred.var = c("age"), grid.resolution = 10) |> autoplot()
p2 = pdp::partial(mars_fit, pred.var = c("bmi"), grid.resolution = 10) |> autoplot()
p3 = pdp::partial(mars_fit, pred.var = c("time"), grid.resolution = 10) |> autoplot()

p4 = pdp::partial(mars_fit, pred.var = c("bmi", "time"),
grid.resolution = 10) |>
pdp::plotPartial(levelplot = FALSE, zlab = "yhat", drape = TRUE, screen = list (z = 40, x = 40))

gridExtra::grid.arrange(p1, p2, p3, ncol =3)
```


# Compare models:
```{r compare}
resamp = resamples(list(PLS =pls_fit,
                        GAM = gam_fit, 
                        MARS = mars_fit))
summary(resamp)

bwplot(resamp, metric = "RMSE")
```




